---
title: "New Mexico - Race and Country of Origin"
format: 
  gfm: default
  html: 
    toc: true
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(paged.print = FALSE)
```

```{r message=FALSE}
library(tidycensus)
library(tidyverse)
library(tigris)
library(patchwork)
options(tigris_use_cache = TRUE)
```

## Motivation

New Mexico's racial demographics are different from the US overall in a number of ways due to its unique history. Firstly, New Mexico has relatively large Hispanic and Native American populations. Secondly, the breakdown of the Hispanic population is unlike any other, in that a significant portion do not identify with Latin America, but directly with Spain. New Mexicans were not strong supporters of the Mexican revolution, and only about 30 years later, New Mexico became part of the US under the Treaty of Hidalgo in 1948.

```{r}
nm_counties <- counties("NM")
```

### Variables of interest

2023

Place of Birth

```         
Total B05002_001
Native B05002_002
Foreign Born B5002_013
```

Of Foreign born

```         
Total B05006_001
Europe B05006_002
Asia B05006_047
Eastern Asia B05006_048
China B05006_049
South Central Asia B05006_056
South Eastern Asia B05006_068
Vietnam B05006_007
Western Asia B05006_079
Africa B05006_095
Oceania B05006_130
Latin America B05006_139
Central America B05006_154
Mexico B05006_160
South America B05006_164
Canada B05006_177
```

Race

These variables include combinations of races

```         
White  B02008_001
Black  B02009_001
NativeAmerican  B02010_001
Asian  B02011_001

Hispanic B03001_003
HispanicMexican B03001_004
HispanicCA B03001_008
```

Measures

```         
Per capita income (inf adj) B19301_001
Median income B19326_001
Gini B19083_001
```

```{r}
pob_vars = c(
  Native = "B05002_002",
  ForeignBorn = "B05002_013"
)
```

```{r}
native_foreign <- get_acs(
  geography = "county",
  state = "NM",
  variables = pob_vars,
  summary_var = "B05002_001",
  year = 2023,
  cache_table = TRUE,
)
```

```{r}
native_foreign <- native_foreign |> left_join(
  nm_counties,
  by = "GEOID"
)  |> select(GEOID, variable, estimate, moe,
            county = NAME.y, summary_est, 
            geometry)
```

```{r}
ggplot(native_foreign, aes(x = variable, y = estimate, fill = county)) +
  geom_col() +
  labs(x = element_blank(),
       y = "Population",
       fill = "County",
       title = "NM Native vs Foreign-born Population",
       subtitle = "By County",
       caption = "acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  theme_bw()
```

```{r}
native_foreign_us <- get_acs(
  geography = "us",
  variables = pob_vars,
  summary_var = "B05002_001",
  year = 2023,
  cache_table = TRUE,
)
```

```{r}
ggplot(native_foreign_us, aes(x = variable, y = estimate)) +
  geom_col(fill = "navy") +
  labs(x = element_blank(),
       y = "Population",
       title = "US Native vs Foreign-born Population",
       subtitle = "By County",
       caption = "acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  theme_bw()
```

```{r}
orig_vars = c(
  Europe = "B05006_002",
  Asia = "B05006_047",
  Eastern_Asia = "B05006_048",
  China = "B05006_049",
  SC_Asia = "B05006_056",
  SE_Asia = "B05006_068",
  Vietnam = "B05006_007",
  Western_Asia = "B05006_079",
  Africa = "B05006_095",
  Oceania = "B05006_130",
  Latin_America = "B05006_139",
  Central_America = "B05006_154",
  Mexico = "B05006_160",
  South_America = "B05006_164",
  Canada = "B05006_177"
)
```

```{r}
foreign_origin <- get_acs(
  geography = "county",
  state = "NM",
  variables = orig_vars,
  summary_var = "B05006_001",
  year = 2023,
  cache_table = TRUE,
)
```

```{r}
foreign_origin <- foreign_origin |> left_join(
  nm_counties,
  by = "GEOID"
)  |> select(GEOID, variable, estimate, moe,
            county = NAME.y, summary_est, 
            geometry)
```

```{r}
continents <- c(
  "Europe",
  "Central_America",
  "Asia",
  "South_America",
  "Africa",
  "Oceania",
  "Canada"
)
```

```{r fig.height=5}
foreign_origin |> 
  filter(variable %in% continents) |> 
  mutate(variable = fct_reorder(variable, desc(estimate))) |> 
  ggplot(aes(x = variable, y = estimate, fill = county)) +
  geom_col() +
  labs(x = "Continent",
       y = "Population",
       fill = "County",
       title = "Origins of foreign-born New Mexico population",
       subtitle = "By Continent",
       caption = "acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30))
```

```{r}
counties_top_8 <- foreign_origin |> 
  group_by(county) |> 
  summarise(est_sum = sum(estimate)) |> 
  slice_max(order_by = est_sum, n = 8) |>
  ungroup()
```

```{r}
counties_top_8
```

```{r}
foreign_origin |> 
  filter(variable %in% continents & county %in% counties_top_8$county) |> 
  mutate(county = fct_reorder(county, desc(summary_est))) |> 
  ggplot(aes(x = county, y = estimate, fill = variable)) +
  geom_col() +
  labs(x = "County",
       y = "Population",
       fill = "Continent",
       title = "Origins of foreign-born New Mexico population",
       subtitle = "Top 10 Counties by Population") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("Africa", "Asia", "Canada", "Central America",
                               "Europe", "Oceania", "South America")) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30))
```

```{r}
race_vars = c(
  White = "B02008_001",
  Black = "B02009_001",
  Native_American = "B02010_001",
  Asian = "B02011_001"
)

races <- get_acs(
  geography = "county",
  state = "NM",
  variables = race_vars,
  year = 2023,
  cache_table = TRUE) |> 
  left_join(
    nm_counties,
    by = "GEOID")  |> 
  select(GEOID, variable, estimate, moe,
            county = NAME.y,geometry)
```

```{r}
races |> 
  filter(county %in% counties_top_8$county) |> 
  mutate(county = fct_reorder(county, desc(estimate))) |> 
  ggplot(aes(x = county, y = estimate, fill = variable)) +
  geom_col() +
  labs(x = "County",
       y = "Population",
       fill = "Race",
       title = "New Mexico population by race",
       subtitle = "Top 10 Counties by Population",
       caption = "Source: census.gov acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("Asian", "Black",
                                  "Native American", "White")) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30))
```

```{r}
hisp_vars = c(
  Non_Hispanic = "B03001_002",
  Hispanic = "B03001_003",
  Hispanic_Mexican  = "B03001_004",
  Hispanic_CA = "B03001_008",
  Hispanic_Spaniard = "B03001_028",
  Hispanic_Spanish = "B03001_029",
  Hispanic_Spanish_American = "B03001_030",
  Hispanic_White = "B03002_013"
)

hispanic <- get_acs(
  geography = "county",
  state = "NM",
  variables = hisp_vars,
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  left_join(nm_counties,
            by = "GEOID") |> 
  select(GEOID, variable, estimate, moe,
            county = NAME.y, summary_est, 
            geometry)


hisp <- hispanic |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rowwise() |> 
  mutate(Hispanic_Spain = sum(c_across(Hispanic_Spaniard:Hispanic_Spanish_American), na.rm = TRUE)) |> 
  select(-c("Hispanic_Spaniard", "Hispanic_Spanish", "Hispanic_Spanish_American")) |> 
  pivot_longer(cols = Non_Hispanic:Hispanic_Spain, names_to = "variable", values_to = "estimate", values_drop_na = T) |> 
  group_by(county, variable) |> 
  mutate(pop = sum(estimate)) |> 
  distinct(county, variable, .keep_all = TRUE)
```

```{r}
hisp_us <- get_acs(
  geography = "us",
  variables = hisp_vars,
  summary_var = "B03001_001",
  year = 2023,
  cache_table = TRUE) |> 
  pivot_wider(names_from = variable, values_from = estimate) |> 
  rowwise() |> 
  mutate(Hispanic_Spain = sum(c_across(Hispanic_Spaniard:Hispanic_Spanish_American), na.rm = TRUE)) |> 
  select(-c("Hispanic_Spaniard", "Hispanic_Spanish", "Hispanic_Spanish_American")) |> 
  pivot_longer(cols = Non_Hispanic:Hispanic_Spain, names_to = "variable", values_to = "estimate", values_drop_na = T)|> 
  group_by(variable) |> 
  mutate(pop = sum(estimate)) |> 
  distinct(variable, .keep_all = TRUE) |> 
  mutate(percent = pop / summary_est)
hisp_us
```

```{r}
hisp |> 
  filter(county %in% counties_top_8$county & !(variable %in% c("Hispanic", "Non_Hispanic"))) |> 
  mutate(county = fct_reorder(county, desc(pop))) |> 
  ggplot(aes(x = county, y = pop, fill = factor(variable, levels = c("Hispanic_White", "Hispanic_Mexican", "Hispanic_Spain", "Hispanic_CA")))) +
  geom_col(position = "dodge") +
  labs(x = "County",
       y = "Population",
       fill = "Race",
       title = "Racial Identification of Hispanic Population",
       subtitle = "Top 10 Counties by Population",
       caption = "Source: census.gov acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("White", "Mexicam", "Spanish", "Central American")) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30))
```

```{r}
hispanic |> 
  filter(county %in% counties_top_8$county & variable %in% c("Hispanic", "Non_Hispanic")) |> 
  mutate(county = fct_reorder(county, desc(estimate))) |> 
  ggplot(aes(x = county, y = estimate, fill = variable)) +
  geom_col(position = "dodge") +
  labs(x = "County",
       y = "Population",
       fill = "Continent",
       title = "Hispanic vs Non-Hispanic population in New Mexico",
       subtitle = "Top 10 Counties by Population",
       caption = "Source: census.gov acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("Hispanic", "Non Hispanic")) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30))
```

```{r}
hisp_us
```

```{r}
hisp_us |> 
  filter(variable %in% c("Hispanic", "Non_Hispanic")) |> 
  ggplot(aes(x = variable, y = estimate)) +
  geom_col(fill = "cyan", color = "black") +
  labs(x = "County",
       y = "Population",
       title = "Hispanic vs Non-Hispanic population in the US",,
       caption = "Source: census.gov acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("Hispanic", "Non Hispanic")) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30))
```

```{r}
measure_vars = c(
  PC_Income = "B19301_001",
  Med_Income = "B19326_001",
  Gini = "B19083_001"
)

measures <- get_acs(
  geography = "county",
  state = "NM",
  variables = measure_vars,
  year = 2023,
  cache_table = TRUE,
) |> 
  left_join(nm_counties,
            by = "GEOID") |> 
  select(GEOID, variable, estimate, moe,
            county = NAME.y, geometry)


hisp |>
  group_by(county) |> 
  select(county, summary_est) |> 
  unique() |> 
  right_join(measures, by = "county") |> 
  filter(county %in% counties_top_8$county & variable == "Gini") |> 
  mutate(county = fct_reorder(county, desc(summary_est))) |> 
  select(c(county, estimate)) |> 
  rename(gini = estimate)
```

```{r}
hisp |>
  group_by(county) |> 
  select(county, summary_est) |> 
  unique() |> 
  right_join(measures, by = "county") |> 
  filter(county %in% counties_top_8$county & variable %in% c("PC_Income", "Med_Income")) |> 
  mutate(county = fct_reorder(county, desc(summary_est))) |> 
  ggplot(aes(x = county, y = estimate, fill = variable)) +
  geom_col(position = "dodge") +
  labs(x = "County",
       y = "Gini Index",
       title = "New Mexico Income",
       subtitle = "Top 10 Counties by Population",
       caption = "Source: census.gov acs5 2023") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_d(labels = c("Median income", "Per capita income")) +
  theme_gray() +
  theme(axis.text.x = element_text(angle = 30))
```

```{r eval=FALSE}
v23 <- load_variables(2023, "acs5", cache = TRUE)
View(v23)
```
